---
title: 'Longitudinal Bulk RNA-Seq Analysis of Canine Lymphoma Samples'
author: 'Elise Hachfeld'
date: '2026-09-08'
output: 
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, message = FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE)

library(here)
library(conflicted)
library(tidyverse)
library(stringr)
library(kableExtra)
library(rtracklayer)
library(tximport)
library(DESeq2)
library(Glimma)
library(apeglm)
library(msigdbr)
library(clusterProfiler)
library(DEGreport)
library(tidytext)
library(org.Hs.eg.db)
library(patchwork)
library(viridis)

conflict_prefer('select', 'clusterProfiler')

project_dir <- 'D:/canine-seq'

walk(c('io.R',
       'counts.R',
       'model.R',
       'annotation.R',
       'enrich.R'),
  ~ source(file.path(project_dir, 'scripts', .x)))
```

```{r include = FALSE}
format_table <- function(x, col.names = NULL, collapse = NULL, digits = 2, caption = NULL){

  tbl <- x |>
    kbl(col.names = col.names, row.names = FALSE, digits = digits, format.args = list(big.mark = ','), caption = paste0('<center><strong>', caption, '</strong></center>'))

  if (!is.null(collapse)) {
    tbl <- tbl |>
      collapse_rows(
        columns = collapse,
        valign = 'top')
    }

  tbl |>
    kable_styling(
      bootstrap_options =
        c('striped', 'hover', 'condensed'),
      fixed_thead = TRUE) |>
    scroll_box(width = '100%', height = '50%')
}
```

# Overview

This report presents the analysis of longitudinal bulk mRNA sequencing data from lymph node fine-needle aspirate samples collected from two dogs with lymphoma enrolled in the ORBIT study. Samples were obtained at baseline (day 0) and at multiple post-treatment time points following administration of Vesicular Stomatitis Virus (VSV-INFb-NIS). The primary goal is to identify genes and biological pathways whose expression changes over time after treatment while accounting for baseline differences between subjects. Due to the small sample size (n = 2), this analysis is exploratory in nature. Additionally, this analysis does not include the examination of genes associated with survival, breed, age, or cancer subtype due to lack of available information.  

# Data Processing
Raw sequencing data were received from the Mayo sequencing core in unaligned CRAM format. Quality control of raw reads was performed using FASTQC. Due to significant contamination from Nextera transposase sequences, trimming of adapters and low quality bases was performed using fastp. Reads were then aligned with HISAT2 using a modified canine reference genome, which combined the genome from canFam4 with the Y and mtDNA from canFam3. Quantification was performed at the gene level using FeatureCounts. Reads that failed to align to the canine reference genome were extracted and re-aligned to the Vesiculovirus Indiana genome (GCF_000850045.1) to detect the presence of viral RNA.

```{r echo = FALSE, message = FALSE}
read.table('D:/canine-seq/runs/run_ORBIT_updatedref_trimtest/report/multiqc_data/bowtie2_pe_plot.txt', sep = '\t', header = TRUE, check.names = FALSE) |>
  mutate(Sample = str_remove(Sample, '.FC.*')) |>
  pivot_longer(cols = 2:7,
               names_to = 'type',
               values_to = 'count') |>
  ggplot(aes(x=Sample, y = count, fill = type)) +
  geom_col(position = 'fill') +
  theme_classic() +
  theme(legend.position = 'bottom') +
  scale_fill_viridis_d(name=NULL, option = 'plasma') +
  guides(fill = guide_legend(nrow = 3, byrow = TRUE)) +
  labs(x = NULL,
       y = 'Percentage of reads',
       title = 'HISAT2 alignment quality') +
  coord_flip()
```

The majority of reads mapped uniquely to the reference genome for every sample. 

```{r echo = FALSE, message = FALSE}
read.table('D:/canine-seq/runs/run_ORBIT_updatedref_trimtest/report/multiqc_data/featureCounts_assignment_plot.txt', sep = '\t', header = TRUE, check.names = FALSE) |>
  mutate(Sample = str_remove(Sample, '.FC.*')) |>
  pivot_longer(cols = 2:6,
               names_to = 'type',
               values_to = 'count') |>
  ggplot(aes(x=Sample, y = count, fill = type)) +
  geom_col(position = 'fill') +
  theme_classic() +
  theme(legend.position = 'bottom') +
  guides(fill = guide_legend(nrow = 3, byrow = TRUE)) +
 scale_fill_viridis_d(name=NULL, option = 'plasma') +
  labs(x = NULL,
       y = 'Percentage of Reads',
       title = 'FeatureCounts assignment') +
  coord_flip()
```

The majority of reads were assigned to a feature. However, some reads were not assigned due to multimapping or lack of features present in the annotation file used.

### Parameters

```{r include=FALSE, message = FALSE, echo = FALSE}
params <- list(min_samples = 2,
               min_count = 10,
               min_alpha = 0.05,
               min_logfold = 0.5)

paths <- list(canine_counts = here('runs',
                            'run_ORBIT_updatedref_trimtest',
                            'results',
                            'featurecounts',
                            'counts.txt'),

              viral_counts = here('runs',
                                  'run_ORBIT_updatedref_trimtest',
                                  'virus',
                                  'featurecounts',
                                  'counts_viral.txt'),

              sample_sheet = here('data',
                                  'ORBIT',
                                  'sample_sheet.csv'),

              gtf = here('data',
                         'reference',
                         'Canis_lupus_familiarisgsd_UU_Cfam_GSD_1.0_Custom',
                         'genes.gtf.gz'))
```

- `min_samples and min_count` = `r params$min_count`: In order to be included, genes must detected at a count greater than or equal to than min_count in at least min_samples. 
- `min_alpha` = `r params$min_alpha`: Adjusted p-value threshold used to define statistical significance.
- `min_logfold` = `r params$min_logfold`: Absolute log2 fold-change threshold used to define meaningful changes in gene expression.

### Data Import

```{r echo = FALSE, message = TRUE}
sample_sheet <- prep_samples(paths$sample_sheet)
count_matrix <- import_feature_counts(paths$canine_counts, sample_ids = sample_sheet$sample_id)
count_df <- count_matrix_to_df(count_matrix, sample_sheet)
tx2gene <- read_tx2gene(paths$gtf)
```

The table below summarizes the samples included in this analysis, including subjects, time, batch, and clinical information.

```{r echo = FALSE, message = FALSE}
sample_sheet |>
  dplyr::select(subject, time, batch, sex, breed, age, cancer_type) |>
  format_table(col.names = c('Subject', 'Time', 'Batch', 'Sex', 'Breed', 'Age', 'Cancer Type'), collapse = 1, caption = 'Sample Overview')
```

# Differential Expression

Differential expression analysis was performed with DESeq2 to identify genes whose expression changed during treatment while accounting for baseline differences between individual dogs.

A likelihood ratio test (LRT) was used to evaluate whether gene expression varies across time points. 
The full model `~ subject + time` includes both subject-specific effects and temporal effects. A reduced model containing only subject effects `~ subject` was compared against the full model. Genes were filtered out if they didn't have a count above `r params$min_count` in `r params$min_samples`  or more samples.

```{r echo = FALSE, message = FALSE}
dds <- create_dds(count_matrix, sample_sheet, design = ~ subject + time, source = 'matrix')
dds <- run_model(dds, params$min_count, params$min_samples, test = 'LRT', reduced = ~ subject)
``` 

### Results

Genes with significant improvement in model fit after inclusion of the time term are considered temporally regulated. Specific coefficient contrasts were examined to identify genes that were differentially expressed at each time point. Fold change values were shrunk using the apeglm shrinkage method. 

```{r echo = FALSE, message = FALSE}
counts <- merge_counts(count_matrix, dds, sample_sheet)

df <- extract_results(dds, params$min_alpha, params$min_logfold) 
  
sym_to_entrez <- get_entrez_ids(unique(df$gene_id), from_type = 'SYMBOL')
#ens_to_entrez <- get_entrez_ids(unique(df$gene_id), from_type = 'ENSEMBL')

df <- df |>
  left_join(sym_to_entrez, by = 'gene_id')
```

```{r echo = FALSE, message = FALSE}
time_coefs <- resultsNames(dds) |>
  str_subset('^time')

subject_coefs <- resultsNames(dds) |>
  str_subset('^subject')

df_sig <- df |>
  dplyr::filter(sig == TRUE,
                shrunk == 'yes')
df_time <- df_sig |>
  dplyr::filter(coefficient %in% time_coefs)

df_subject <- df_sig |>
  dplyr::filter(coefficient %in% subject_coefs)

df_up <- df_time |>
  dplyr::filter(regulation == 'downregulated')

df_down <- df_time |>
  dplyr::filter(regulation == 'downregulated')
```

Differential expression results are explored through interactive MA and volcano plots, which allow visualization of effect sizes and p-values. 

```{r echo = FALSE, message = TRUE}
glimmaMA(dds, groups = colData(dds)$time)
```

```{r echo = FALSE, message = TRUE}
glimmaVolcano(dds, groups = colData(dds)$time)
```

```{r echo = FALSE, message = FALSE}
df_time |>
  dplyr::filter(abs(log2FoldChange) > 1) |>
  dplyr::select(regulation, coefficient, gene_id, padj, log2FoldChange) |>
  arrange(regulation, coefficient, desc(abs(log2FoldChange)), padj) |>
  format_table(col.names = c('Regulation', 'Coefficient',  'Gene Name', 'Adjusted p-value', 'Log2 Fold-Change (shrunk)'), collapse = c(1,2), digits = 6, caption = 'Differentially expressed genes (abs(log2FC) > 1)')
```

### Candidate Genes of Interest
```{r echo = FALSE, message = FALSE}
genes_of_interest <- unique(df_time$gene_id)

plots <- lapply(genes_of_interest, \(g) plot_gene_expression(counts, g))

names(plots) <- genes_of_interest

genes_to_plot <- c('TNF', 'FOSB', 'EGR1', 'EGR2', 'EGR3', 'RETN', 'TLR4', 'CD3E', 'CD34', 'ADRB2', 'FGF23', 'CXCL12', 'JCHAIN', 'TCF7', 'ZAP70', 'SOX13', 'LEF1')

wrap_plots(plots[genes_to_plot],
           ncol = 3,
           guides = 'collect',
           axes_titles = 'collect',
           axes = 'collect')
```

# Quality Control 

### Count Distribution

The distribution of raw gene counts for each sample is visualized below. RNA-seq count data are typically right-skewed, with most genes exhibiting low or no expression and a minority of highly expressed genes. There should be similar overall distributions between samples, but some variability is expected.

```{r echo = FALSE}
count_df |>
  ggplot() +
  geom_histogram(aes(x = log2(count + 1)), position = 'identity', bins = 200) +
    facet_grid(rows = vars(subject),
             cols = vars(time), labeller = label_both) +
  theme_classic() +
  labs(x = expression(log[2] (count + 1)),
       y = 'Gene frequency',
       title = 'Distribution of raw counts')
```

```{r echo = FALSE, message = FALSE}
counts |>
  group_by(subject, time, normalized) |>
   dplyr::summarize(total_count = sum(count, na.rm =TRUE),
                    mean_count = mean(count, na.rm =TRUE),
                    sd_count = sd(count, na.rm =TRUE),
                    max_count = max(count, na.rm =TRUE),
                    num_genes = sum(count > 0, na.rm =TRUE))|>
  arrange(subject, time) |>
  dplyr::filter(normalized == 'yes') |>
  dplyr::select(-normalized) |>
  format_table(col.names = c('Subject', 'Time', 'Total Count', 'Mean Count', 'Standard Deviation', 'Max Count', 'Number of Genes'), digits = 0, collapse = c(1, 2), caption = 'Normalized counts')
```

### Sample Relationships

A multidimensional scaling plot was used to visualize relationships between samples. Samples clustering by subject indicate differences in expression between dogs, while clustering by time point may suggest changes in expression because of the treatment. 

```{r echo = FALSE, message = FALSE}
glimmaMDS(dds)
```

The first dimensions of the MDS plot indicate that differences between subjects account for much of the variation in the data. Given this, subject was included as a variable in subsequent modeling. 

### Dispersion Estimation

DESeq2 models the relationship between gene expression mean and variability using gene-wise dispersion estimates. The black points represent dispersion estimates for each gene. DESeq2 adjusts these estimates based on the fitted curve (red), resulting in shrunken dispersion estimates (blue). Systematic deviations of the dispersion estimates from the fitted curve may indicate outliers.

```{r echo = FALSE, message = FALSE}
plotDispEsts(dds)
```

### Log Fold-Change Shrinkage

DESeq2 estimates log2 Fold-Change values for each gene to describe the effect size. However, these estimates can be inflated in genes with small counts. The apeglm shrinkage method reduces noise in fold-change estimates while preserving large biologically meaningful effects.

```{r echo = FALSE, message = FALSE}
res <- results(dds)
resLFC <- lfcShrink(dds, coef = 'time_49_vs_0', type = 'apeglm')

par(mfrow = c(1, 2))  # 1 row, 2 columns
plotMA(res, alpha = 0.1, ylim = c(-2, 2), cex = 1.5, main = 'MA-plot before LFC shrinkage')
plotMA(resLFC, alpha = 0.1, ylim = c(-2, 2), cex = 1.5, main = 'MA-plot after LFC shrinkage')
```

```{r echo = FALSE}
df |>
  group_by(coefficient, shrunk) |>
  summarise(DE_genes = sum(sig),
            Up = sum(regulation == 'upregulated'),
            Down = sum(regulation == 'downregulated'),
            .groups = 'drop_last') |>
  arrange(coefficient) |>
  format_table(col.names = c('Coefficient', 'Shrunk?', 'Number of DE Genes', 'Upregulated', 'Downregulated'), collapse = 1, caption = 'Effect of fold change shrinkage on DE genes')
```


Shrinkage of fold change values resulted in a smaller number of differentially expressed genes being identified, however, we can be more confident that those that were identified were accurate. 

# Functional Enrichment Analysis
## Over-Representation Analysis
```{r eval = FALSE}
df_up_filter <- df |>
  dplyr::filter(is.na

df_universe <- df |>
  dplyr::filter(!is.na(entrez_id),
                shrunk == 'yes',
                coefficient != 'subject_2_vs_1')|>
  dplyr::pull(entrez_id)

ego_up <- enrichGO(gene=df_up$entrez_id,
                OrgDb= org.Hs.eg.db,
                keyType = 'ENTREZID',
                ont= 'BP',
                universe=df_universe,
                readable = TRUE)

s_ego_up <-clusterProfiler::simplify(ego_up)

heatplot(ego_up, showCategory=10)
#goplot(ego_up)

dotplot(s_ego_up,showCategory=20,font.size=10,label_format=70)+
  scale_size_continuous(range=c(1, 7))+
  theme_minimal() +
  ggtitle('GO Enrichment of up-regulated genes')


ego_up_df <- ego_up@result |>
  as.data.frame()

ego_up_df |>
  dplyr::filter(p.adjust < 0.05) |>
  arrange(p.adjust, desc(FoldEnrichment)) |>
  select(Description, GeneRatio, geneID, FoldEnrichment, p.adjust) |>
  format_table(col.names = c('Pathway', 'Number of Genes', 'Enriched Genes', 'Fold Enrichment', 'Adjusted p-value'))
```

```{r eval = FALSE}
df_filter_down <- df_time |>
  dplyr::filter(regulation == 'downregulated',
                !is.na(entrez_id))

ego_down <- enrichGO(gene= df_filter_down$entrez_id,
                OrgDb= org.Hs.eg.db,
                keyType = 'ENTREZID',
                ont= 'BP',
                universe=df_universe,
                readable = TRUE)

s_ego_down <-clusterProfiler::simplify(ego_down)

heatplot(ego_down, showCategory=10)
#goplot(ego_down)

dotplot(s_ego_down,showCategory=20,font.size=10,label_format=70)+
  scale_size_continuous(range=c(1, 7))+
  theme_minimal() +
  ggtitle('GO Enrichment of down-regulated genes')
```

```{r eval = FALSE}
for (coef in resultsNames(dds)[-c(1, 2)]){
  print(coef)
  
  ora_df <- df_time |>
    dplyr::filter(regulation == 'upregulated',
                  coefficient == coef,
                  !is.na(entrez_id))
  
  universe <- df |>
    dplyr::filter(shrunk == 'yes',
                  coefficient == coef,
                  !is.na(entrez_id)) |>
    dplyr::pull(entrez_id)
  
  ego <- enrichGO(gene=ora_df$entrez_id,
                OrgDb= org.Hs.eg.db,
                keyType = 'ENTREZID',
                ont= 'BP',
                universe=universe,
                readable = TRUE)
  
  s_ego <-clusterProfiler::simplify(ego)
  
  print(head(s_ego)) 
  if (nrow(s_ego) > 0){
    plot <- dotplot(s_ego,showCategory=20,font.size=10,label_format=70)+
  scale_size_continuous(range=c(1, 7))+
  theme_minimal() +
  ggtitle('GO Enrichment of up-regulated genes')
  print(plot)
  }
}
```

```{r eval = FALSE}
 for (coef in resultsNames(dds)[-c(1, 2)]){
  print(coef)
  
  ora_df <- df_time |>
    dplyr::filter(regulation == 'downregulated',
                  coefficient == coef,
                  !is.na(entrez_id))
  
  universe <- df |>
    dplyr::filter(shrunk == 'yes',
                  coefficient == coef,
                  !is.na(entrez_id)) |>
    dplyr::pull(entrez_id)
  
  ego <- enrichGO(gene=ora_df$entrez_id,
                OrgDb= org.Hs.eg.db,
                keyType = 'ENTREZID',
                ont= 'BP',
                universe=universe,
                readable = TRUE)
  
  s_ego <-clusterProfiler::simplify(ego)
  
  print(head(s_ego)) 
  if (nrow(s_ego) > 0){ 
    plot <- dotplot(s_ego,showCategory=20,font.size=10,label_format=70)+
    scale_size_continuous(range=c(1, 7))+
    theme_minimal() +
    ggtitle('GO Enrichment of down-regulated genes')
    print(plot)
  }
 }
```

## Gene Set Enrichment

```{r eval = FALSE}
df_shrunk <- df |>
  dplyr::filter(shrunk == 'yes')

df_hallmark <- run_gsea(dds, df_shrunk, collection = 'H', match_by = 'gene_symbol')
df_kegg <- run_gsea(dds, df_shrunk, collection = 'C2', subcollection = 'CP:KEGG_MEDICUS', match_by = 'gene_symbol')
df_onco_sig <- run_gsea(dds, df_shrunk, collection = 'C6', match_by = 'gene_symbol')
```

```{r eval = FALSE}
df_kegg |>
  dplyr::filter(coefficient != 'subject_2_vs_1') |>
  dplyr::mutate(pathway = str_remove(pathway, 'Kegg Medicus')) |>
  dplyr::select(regulation, coefficient, pathway, core_enrichment, NES, p.adjust) |>
  arrange(regulation, coefficient, p.adjust, desc(abs(NES))) |>
  format_table(col.names = c('Regulation', 'Coefficient',  'Pathway', 'Core Enrichment', 'Normalized Enrichment Score', 'Adjusted p-value', collapse = c(1,2))
```

```{r eval = FALSE}
kegg_plots <- plot_gsea(df_kegg, min_alpha=10, n = 3)
(kegg_plots[[1]] / kegg_plots[[2]]) + plot_layout(axes = 'collect', axis_titles = 'collect')

hallmark_plots <- plot_gsea(df_hallmark, min_alpha=10, n = 3)
(hallmark_plots[[1]] / hallmark_plots[[2]]) + plot_layout(axes = 'collect', axis_titles = 'collect')

onco_sig_plots <- plot_gsea(df_onco_sig, min_alpha=10, n = 3)
(onco_sig_plots[[1]] / onco_sig_plots[[2]]) + plot_layout(axes = 'collect', axis_titles = 'collect')
```

# Clustering  (in progress)
```{r eval = FALSE}
library(dendextend)
test <- counts |>
  dplyr::filter(normalized == 'yes') |>
  dplyr::select(gene_id, count, subject, time) |>
  dplyr::filter(gene_id %in% df_time$gene_id) |>
  group_by(gene_id, time) |>
  summarize(mean_count = mean(count)) |>
  pivot_wider(names_from = gene_id,
              values_from = mean_count)

test.cor <- dplyr::select(test, -time) |>
  cor(use='pairwise.complete.obs')

test.dist <- as.dist(1 - test.cor)
test.tree <- hclust(test.dist, method='complete')

dend <- as.dendrogram(test.tree)
clusters <- cutree(dend, k=4)

plot(color_branches(dend, k=6),leaflab='none')


clusters <- cutree(dend, k=3)
clusters.df <- data.frame(gene = names(clusters), cluster = clusters)
```

```{r eval = FALSE}
vst <- vst(dds, blind = TRUE)
vst_mat <- assay(vst)
cluster_vst <- vst_mat[df_time$gene_id, ]
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across sample groups

clusters <- degPatterns(cluster_vst, metadata = sample_sheet, time = 'time')
#clusters <- degPatterns(cluster_rlog, metadata = sample_sheet, time = 'time', col = 'subject')

cluster_df <- clusters$df 
```

# Viral RNA Detection (in progress)
```{r eval = FALSE}
viral_counts <- read.table(paths$viral_counts, sep = '\t', header = TRUE, check.names = FALSE) |>
  tibble::column_to_rownames('Geneid') |>
  dplyr::select(-(1:5)) |>
  dplyr::rename_with(~ str_remove_all(.x, 'runs\\/.*sort\\/|_aligned_sorted_viral.bam')) 

viral_df <- count_matrix_to_df(viral_counts, sample_sheet)
viral_df |>
  ggplot(aes(x=time, y=count, color = gene_id)) +
  geom_boxplot() +
  geom_point(width = 0, size = 1, color = 'black', alpha = 0.5) +
  facet_wrap(~gene_id) +
  theme_classic()
```

# Future Analyses/TO-DO
1. Cell Type Deconvolution
2. TCR/BCR Investigation



```{r}
# add this plot for different pathways
counts |>
  dplyr::filter(gene_id %in% unique(df_time$gene_id) & normalized == 'yes') |>
  ggplot(aes(x=time, y = count)) +
  geom_boxplot() +
  facet_wrap(~gene_id, scales = 'free')
```


# Key Findings
**write this**

# Session Information
```{r}
sessionInfo()
```