---
title: "tximport"
output: html_document
date: "2026-08-19"
editor_options: 
  chunk_output_type: console
---

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

#BiocManager::install("Glimma")
#BiocManager::install("DESeq2")
#BiocManager::install("GenomicRanges")
#BiocManager::install("rtracklayer")
#install.packages("remotes", repos='http://cran.us.r-project.org')
#remotes::install_github("AllenInstitute/GeneOrthology")

library(tidyverse)
library(tximport)
library(DESeq2)
library(Glimma)
library(rtracklayer)
```

```{r}
# SET WORKING DIRECTORY
setwd("D:/canine-seq")
wd <- getwd()

# SET CONSTANTS
min_alpha = 0.05
min_logfold = 0.5
time_levels = c(0, 3, 7, 10, 14, 28, 49)
min_count = 10
min_samples = 2

# SET FILE PATHS
sample_sheet_path = "D:/canine-seq/data/ORBIT/sample_sheet.csv"
gtf_file_path = "D:/canine-seq/reference/dog/canfam4/Canis_lupus_familiarisgsd.UU_Cfam_GSD_1.0.116.gtf.gz"
kallisto_path = "D:/canine-seq/results/ORBIT/kallisto"
```

```{r}
build_tx2gene <- function(gtf_file_path=gtf_file_path){
  tx2gene <- rtracklayer::readGFF(gtf_file_path) |>
    as.data.frame() |>
    dplyr::select(transcript_id, gene_id, gene_name) |>
    distinct() |>
    dplyr::filter(!is.na(transcript_id)) |>
    mutate(gene_name = coalesce(gene_name, gene_id))
    return(tx2gene)}

prepare_samples <- function(sample_sheet_path=sample_sheet_path, time_levels=time_levels){
  sample_sheet <- read.csv(sample_sheet_path) |>
    mutate(time = factor(timepoint_days, levels = time_levels),
           time_num = as.numeric(timepoint_days),
           subject = factor(subject),
           sample_name = paste0("Subject ", subject, ", ", "Day ", time)) |>
    select(-timepoint_days)
  return(sample_sheet)}

import_kallisto <- function(kallisto_path=kallisto_path){
  files <- file.path(list.dirs(kallisto_path)[-1], "abundance.h5")
  names(files) <- gsub(paste0(kallisto_path, "/"), "", list.dirs(kallisto_path)[-1])
  txi <- tximport(files, type = "kallisto",
                tx2gene = tx2gene,
                txOut = FALSE,
                ignoreTxVersion = TRUE)}


run_model <- function(txi=txi,
                      sample_sheet=sample_sheet,
                      design,
                      reduced,
                      test = "Wald",
                      col,
                      min_count = min_count,
                      min_samples = min_samples) {
  dds <- DESeqDataSetFromTximport(txi,
                                  colData = sample_sheet,
                                  design = design)
  
  keep <- rowSums(counts(dds) >= min_count) >= min_samples
  dds <- dds[keep,]

  if (test == "LRT") {
    dds <- DESeq(dds, test = test, reduced = reduced)
  }
  else {
    dds <- DESeq(dds, test = test)
  }
  return(dds)}
  
process_results <- function(dds=dds, tx2gene=tx2gene, alpha = 0.05){
  res <- results(dds)
  print(summary(res))
  res <- res |>
    as.data.frame() |>
    rownames_to_column("gene_id") |>
    left_join(tx2gene, by = "gene_id") |>
    select(-transcript_id) |>
    distinct(gene_id, .keep_all = TRUE) |>
    filter(!is.na(padj)) |>
    arrange(padj) |>
    filter(padj < alpha)
    return(res)}

process_counts <- function(counts) {
  counts <- counts |>
  t()}
```


```{r}
# IMPORT NECESSARY DATA
tx2gene <- build_tx2gene(gtf_file_path)
sample_sheet <- prepare_samples(sample_sheet_path, time_levels)
txi <- import_kallisto(kallisto_path)

counts <- txi$counts


geom_point()
```

```{r}
#QUALITY CONTROL
# a plot with the sum of counts for each sample



```


```{r}
# MODEL FITTING

# TIME + SUBJECT LRT W/ TIME AS FACTOR # non linear effects over time
# shows the genes that are differentially expressed over time accounting for variation between subjects
dds <- run_model(txi,
                  sample_sheet,
                  design = ~ subject + time,
                  reduced = ~ subject,
                  test = "LRT",
                  col = time,
                  min_count = min_count,
                  min_samples = min_samples)


plotPCA(rlog(dds), intgroup = c("subject"))


glimmaMA(dds, groups = colData(dds)$time)
res <- process_results(dds, tx2gene, min_alpha)
```

```

```{r}
# examples for one sample
summary(raw_counts[,1])
#hist(log2(raw_counts[,1] + 1))
sum(raw_counts[,1] > 0) #number of expressed genes
top_idx <- order(raw_counts[[1]], decreasing = TRUE)

head(data.frame(gene = rownames(raw_counts)[top_idx],
                count = raw_counts[[1]][top_idx]),20)

ggplot(raw_counts) + 
  geom_histogram(aes(x=raw_counts[[1]]), stat = "bin", bins = 200) + 
  xlab("Raw counts") + 
  ylab("Number of genes")

ggplot(raw_counts) +
  geom_histogram(aes(x = log2(raw_counts[,1] + 1)),bins = 100) +
  xlab("log2(count + 1)") 
```


