---
title: "Longitudinal Bulk RNA-Seq Analysis of Canine Lymphoma Samples"
author: "Elise Hachfeld"
output: 
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
date: "2026-08-19"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

# Overview
add info here

# Setup
## Load Required Packages
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install(c("tidyverse",
                       "stringr",
                       "rtracklayer",
                       "tximport",
                       "DESeq2",
                       "Glimma",
                       "apeglm",
                       "msigdbr",
                       "org.Hs.eg.db",
                       "clusterProfiler"))
library(tidyverse)
library(stringr)
library(rtracklayer)
library(tximport)
library(DESeq2)
library(Glimma)
library(apeglm)
library(msigdbr)
library(clusterProfiler)
library(org.Hs.eg.db)
```

## Helper Functions
### prep_tx2gene
Description:    
**Input:**    
**Output:**     
```{r}
prep_tx2gene <- function(gtf_path) {
  print(paste0("Reading .gtf file from ", gtf_path, "..."))
  tx2gene <- rtracklayer::readGFF(gtf_path) |>
    as.data.frame() |>
    dplyr::select(transcript_id,
                  gene_id,
                  gene_name,
                  gene_biotype) |>
    dplyr::filter(!is.na(transcript_id)) |>
    dplyr::distinct() |>
    dplyr::mutate(gene_name = coalesce(gene_name,
                                       gene_id)) # if no gene name present, use gene id
  print(paste0("Found ", length(unique(tx2gene$transcript_id)), " unique transcripts and ", length(unique(tx2gene$gene_id)), " unique genes."))
  return(tx2gene)
}
```

### prep_samples
Description:    
**Input:**    
**Output:**     
```{r}
prep_samples <- function(sample_sheet_path) {
  print(paste0("Reading .csv file from ", sample_sheet_path, "..."))
  sample_sheet <- read.csv(sample_sheet_path) |>
    dplyr::mutate(time_num = as.numeric(time),
                  time = factor(time, levels = sort(unique(time))),
                  subject = factor(subject, levels = sort(unique(subject))),
                  sample_name = paste0("Subject ", subject, ", ", "Day ", time)) # readable sample name for visualizations
  rownames(sample_sheet) <- sample_sheet$sample_id
  print(paste0("Found ", length(sample_sheet$sample_id), " samples in sample sheet."))
  return(sample_sheet)
}
```
### prep_kallisto
Description:    
**Input:**    
**Output:**     
```{r}
prep_kallisto <- function(kallisto_path,
                          tx2gene,
                          sample_sheet,
                          transcripts = FALSE) {
  # NOTE: this will not include samples in the file path if they are not in the sample sheet
  print(paste0("Reading abundance.h5 files from ", kallisto_path, "..."))
  files <- file.path(kallisto_path, sample_sheet$sample_id, "abundance.h5") 
  print(paste0("Found ", length(files), " files in ", kallisto_path))
  names(files) <- sample_sheet$sample_id
  txi <- tximport(files,
                  type = "kallisto",
                  tx2gene = tx2gene,
                  txOut = transcripts, # change this to true if you want transcript level data
                  ignoreTxVersion = TRUE)
  return(txi)
}
```
### prep_homologs
Description:    
**Input:**    
**Output:**     
```{r}
prep_homologs <- function(homolog_path) {
  print(paste0("Reading csv from ", homolog_path, "..."))
  homologs <- read.csv(homolog_path, sep = "\t") |>
    dplyr::filter(ref_species == "homo_sapiens") |>
    dplyr::select(ref_gene_stable_id, ref_gene_name, query_gene_stable_id, query_gene_name) |>
    dplyr::rename(human_gene_id = ref_gene_stable_id,
                  human_gene_name = ref_gene_name,
                  dog_gene_id = query_gene_stable_id,
                  dog_gene_name = query_gene_name)
  return(homologs)
}
```
### run_model
Description:    
**Input:**    
**Output:**     
```{r}
run_model <- function(txi = txi, sample_sheet, design, reduced, test = "Wald", col, min_count = 10, min_samples = 3) {
  dds <- DESeqDataSetFromTximport(txi, colData = sample_sheet, design = design)
  keep <- rowSums(counts(dds) >= min_count) >= min_samples # filtering for low counts
  dds <- dds[keep, ]

  if (test == "LRT") {
    dds <- DESeq(dds, test = test, reduced = reduced)
  } else {
    dds <- DESeq(dds, test = test)
  }
  return(dds)
}
```
### process_counts
Description:    
**Input:**    
**Output:**     
```{r}
process_counts <- function(counts, sample_sheet) {
  counts <- counts |>
    as.data.frame() |>
    rownames_to_column("gene_id") |>
    pivot_longer(
      cols = starts_with("OR"),
      names_to = "sample_id",
      values_to = "count"
    ) |>
    left_join(sample_sheet, by = "sample_id") |>
    left_join(tx2gene, by = "gene_id") |>
    dplyr::select(-transcript_id) |>
    dplyr::distinct(gene_id, sample_id, .keep_all = TRUE)
  return(counts)
}
```
### process_results
Description:    
**Input:**    
**Output:**     
```{r}
process_results <- function(res, tx2gene, min_alpha = 0.05, min_logfold = 1) {
  res <- res |>
    as.data.frame() |>
    rownames_to_column("gene_id") |>
    left_join(tx2gene, by = "gene_id") |>
    dplyr::select(-transcript_id) |>
    dplyr::distinct(gene_id, .keep_all = TRUE) |>
    dplyr::filter(!is.na(padj)) |> # genes with extreme outliers are set to NA
    dplyr::arrange(padj) |>
    dplyr::filter(padj <= min_alpha) |>
    dplyr::filter(abs(log2FoldChange) >= min_logfold)
  return(res)
}
```
### extract_coefs
Description:    
**Input:**    
**Output:**     
```{r}
extract_coefs <- function(dds, tx2gene, homologs, min_alpha, min_logfold) {
  # getting a data frame that has log fold change, p val and gene name for every coefficient
  coefficients <- resultsNames(dds)

  df_list <- list()
  for (coef in coefficients[-1]) {
    res_shrunk <- lfcShrink(dds, coef = coef, type = "apeglm")
    res_df <- process_results(res_shrunk, tx2gene, min_alpha, min_logfold) |>
      dplyr::mutate(coefficient = coef)
    df_list <- append(df_list, list(res_df))
  }

  df <- do.call(rbind, df_list) |>
    left_join(homologs, by = c("dog_gene_id" = "gene_id"))
  return(df)
}
```

## Set Constants
### Parameters
```{r}
min_alpha <- 0.05
min_logfold <- 0
min_count <- 10
min_samples <- 2
```

### File Locations
```{r}
setwd("D:/canine-seq")
sample_sheet_path <- "D:/canine-seq/data/ORBIT/sample_sheet.csv"
gtf_path <- "D:/canine-seq/reference/dog/canfam4/Canis_lupus_familiarisgsd.UU_Cfam_GSD_1.0.116.gtf.gz"
kallisto_path <- "D:/canine-seq/results/ORBIT/kallisto"
homolog_path <- "D:/canine-seq/data/ORBIT/Canis_lupus_familiaris-GCA_011100685.1-2021_03-homology.tsv"
```

## Data Import
```{r}
tx2gene <- prep_tx2gene(gtf_path)
sample_sheet <- prep_samples(sample_sheet_path)
txi <- prep_kallisto(kallisto_path, tx2gene, sample_sheet)
homologs <- prep_homologs(homolog_path)
counts <- process_counts(txi$counts, sample_sheet)
```


# Quality Control
```{r}
# QUALITY CONTROL
# Count distribution per sample. They should have about equal distributions to fulfill the assumptions of the DESeq2 model.
counts |>
  ggplot() +
  geom_histogram(aes(x = log2(count + 1)), position = "identity", bins = 200) +
  labs(
    x = "log2(count +1)",
    y = "Number of Genes"
  ) +
  facet_grid(
    rows = vars(subject),
    cols = vars(time)
  )
```


# Differential Expression
```{r}
# MODEL FITTING
# TIME + SUBJECT LRT W/ TIME AS FACTOR # non linear effects over time
# shows the genes that are differentially expressed over time accounting for variation between subjects
dds <- run_model(txi,
  sample_sheet,
  design = ~ subject + time,
  reduced = ~subject,
  test = "LRT",
  col = time,
  min_count = min_count,
  min_samples = min_samples
)

counts_norm <- process_counts(counts(dds, normalized = TRUE), sample_sheet)

df_list <- list()
for (coef in resultsNames(dds)[-1]) {
  res_shrunk <- lfcShrink(dds, coef = coef, type = "apeglm")
  res_df <- process_results(res_shrunk, tx2gene, min_alpha = 100000, min_logfold = 0) |>
    mutate(coefficient = coef)
  df_list <- append(df_list, list(res_df))
}

df <- do.call(rbind, df_list)

df <- left_join(df, homologs, by = c("gene_id" = "dog_gene_id"))

df_filter <- df |>
  mutate(regulation = case_when(
    (padj < min_alpha & abs(log2FoldChange) > min_logfold & log2FoldChange > 0) ~ "upregulated",
    (padj < min_alpha & abs(log2FoldChange) > min_logfold & log2FoldChange < 0) ~ "downregulated",
    T ~ "nonDE"
  )) |>
  filter(regulation != "nonDE" & coefficient != "subject_2_vs_1" & coefficient != "time_49_vs_0" & coefficient != "time_10_vs_0") |>
  mutate(time = str_extract(coefficient, "(?<=time_)\\d+"))

# df <- extract_coefs(dds, tx2gene, homologs, min_alpha = 1000000, min_logfold = 0)
```

```{r}
glimmaMDS(dds) # color by subject, shape by time.
# 46% of variance explained by subject

# normalizationFactors(dds)
plotDispEsts(dds) # Black points should cluster around the fitted line.

# visualizing shrinkage of log fold change
res <- results(dds)
plotMA(res, alpha = 0.1, ylim = c(-2, 2), cex = 1.5, main = "MA-plot before LFC shrinkage")

resLFC <- lfcShrink(dds, coef = "time_49_vs_0", type = "apeglm")
plotMA(resLFC, alpha = 0.1, ylim = c(-2, 2), cex = 1.5, main = "MA-plot after LFC shrinkage")

glimmaMA(dds, groups = colData(dds)$time)
glimmaVolcano(dds, groups = colData(dds)$time)

# plotCounts(dds, gene = "ENSCAFG00805014647", intgroup = c("time"), main = "ENSCAFG00805014647")
```

```{r eval = FALSE}
up <- df_filter |>
  filter(regulation == "upregulated") |>
  distinct(gene_id) |>
  pull(gene_id) |>
  as.list()

down <- df_filter |>
  filter(regulation == "downregulated") |>
  distinct(gene_id) |>
  pull(gene_id) |>
  as.list()


counts_norm |>
  filter(gene_id %in% up) |>
  ggplot(aes(x = time_num, y = count, color = subject)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE) +
  theme_bw() +
  facet_wrap(~gene_name, scales = "free") +
  labs(
    x = "Time (days)",
    y = "Normalized count",
    title = "Upregulated Genes"
  )


counts_norm |>
  filter(gene_id %in% down) |>
  ggplot(aes(x = time_num, y = count, color = subject)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE) +
  theme_bw() +
  facet_wrap(~gene_name, scales = "free") +
  labs(
    x = "Time (days)",
    y = "Normalized count",
    title = "Downregulated Genes"
  )
```

# Gene Set Enrichment
```{r}
msigdbr_collections()

hallmark <- msigdbr(collection = "H") |>
  dplyr::select(gs_name, ensembl_gene)
head(hallmark)

h2 <- msigdbr(collection = "H") |>
  dplyr::select(gs_name, gs_description)


C4 <- msigdbr(collection = "C4") |>
  dplyr::select(gs_name, ensembl_gene)
head(C4)
C42 <- msigdbr(collection = "C4") |>
  dplyr::select(gs_name, gs_description)


C7 <- msigdbr(collection = "C7", subcollection = "IMMUNESIGDB") |>
  dplyr::select(gs_name, ensembl_gene)
head(C7)
C72 <- msigdbr(collection = "C7", subcollection = "IMMUNESIGDB") |>
  dplyr::select(gs_name, gs_description)
```

```{r}
time_3 <- df |>
  filter(coefficient == "time_3_vs_0" & !is.na(human_gene_name)) |>
  arrange(padj, log2FoldChange) |>
  pull(log2FoldChange, name = human_gene_id)

eh3 <- GSEA(time_3, TERM2GENE = hallmark, TERM2NAME = h2, seed = TRUE)

time_7 <- df |>
  filter(coefficient == "time_7_vs_0" & !is.na(human_gene_name)) |>
  arrange(padj, log2FoldChange) |>
  pull(log2FoldChange, name = human_gene_id)

eh7 <- GSEA(time_7, TERM2GENE = hallmark, TERM2NAME = h2, seed = TRUE)

time_10 <- df |>
  filter(coefficient == "time_10_vs_0" & !is.na(human_gene_name)) |>
  arrange(padj, log2FoldChange) |>
  pull(log2FoldChange, name = human_gene_id)

eh10 <- GSEA(time_10, TERM2GENE = hallmark, TERM2NAME = h2, seed = TRUE)

time_28 <- df |>
  filter(coefficient == "time_28_vs_0" & !is.na(human_gene_name)) |>
  arrange(padj, log2FoldChange) |>
  pull(log2FoldChange, name = human_gene_id)

eh28 <- GSEA(time_28, TERM2GENE = hallmark, TERM2NAME = h2, seed = TRUE)

time_49 <- df |>
  filter(coefficient == "time_49_vs_0" & !is.na(human_gene_name)) |>
  arrange(padj, log2FoldChange) |>
  pull(log2FoldChange, name = human_gene_id)

eh49 <- GSEA(time_49, TERM2GENE = hallmark, TERM2NAME = h2, seed = TRUE)
```


```{r}
all <- df |>
  filter(!is.na(human_gene_name)) |>
  arrange(padj, log2FoldChange) |>
  pull(log2FoldChange, name = human_gene_id)

ehall <- GSEA(all, TERM2GENE = hallmark, TERM2NAME = h2, seed = TRUE)

# all_C4 <- GSEA(all, TERM2GENE = C4, TERM2NAME = C42, seed=TRUE)
# all_C7 <- GSEA(all, TERM2GENE = C7, TERM2NAME = C72, seed=TRUE)


enrichplot::gseaplot2(ehall, geneSetID = c(1, 2, 3))
```

```{r}
# GSEA_GO <- gseGO(geneList = all,
#                   ont = "BP",
#                   OrgDb = org.Hs.eg.db,
#                   keyType = "ENSEMBL",
#                   seed = TRUE)

df_3 <- as.data.frame(eh3)
df_7 <- as.data.frame(eh7)
df_10 <- as.data.frame(eh10)
df_28 <- as.data.frame(eh28)
df_49 <- as.data.frame(eh49)


result_df <- as.data.frame(ehall)


write.csv(result_df, "result.csv")
```
