---
title: "Longitudinal Bulk RNA-Seq Analysis of Canine Lymphoma Samples"
author: "Elise Hachfeld"
output: 
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
date: "2026-08-19"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE)
setwd("D:/canine-seq")
source("scripts/functions.R")
```

# Overview

add info here

# Setup

## Load Required Packages

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

#BiocManager::install(c("tidyverse",
#                       "stringr",
#                       "kableExtra",
#                       "rtracklayer",
#                       "tximport",
#                       "DESeq2",
#                       "Glimma",
#                       "apeglm",
#                       "msigdbr",
#                       "org.Hs.eg.db",
#                       "clusterProfiler"))

library(tidyverse)
library(stringr)
library(kableExtra)
library(rtracklayer)
library(tximport)
library(DESeq2)
library(Glimma)
library(apeglm)
library(msigdbr)
library(clusterProfiler)
library(org.Hs.eg.db)
```

## Set Constants

### Parameters

```{r}
min_samples <- 2
min_count <- 10
min_alpha <- 0.05
min_logfold <- 0.5
```

### File Locations

```{r}
sample_sheet_path <- "D:/canine-seq/data/ORBIT/sample_sheet.csv"
gtf_path <- "D:/canine-seq/data/reference/Canis_lupus_familiarisgsd_UU_Cfam_GSD_1.0_116/Canis_lupus_familiarisgsd.UU_Cfam_GSD_1.0.116.gtf.gz"
kallisto_path <- "D:/canine-seq/runs/run_ORBIT_canfam4/results/kallisto"
homolog_path <- "D:/canine-seq/data/reference/Canis_lupus_familiarisgsd_UU_Cfam_GSD_1.0_116/Canis_lupus_familiaris-GCA_011100685.1-2021_03-homology.tsv"
mayo_gene_counts <- "D:/mayo_results/merged_gene_counts.txt"
```

## Data Import

```{r}
sample_sheet <- prep_samples(sample_sheet_path)
tx2gene <- prep_tx2gene(gtf_path)
homologs <- prep_homologs(homolog_path)
txi <- prep_kallisto(kallisto_path, tx2gene, sample_sheet)
counts <- process_counts(txi$counts, sample_sheet, tx2gene)
```

```{r echo = FALSE}
sample_sheet |>
  dplyr::select(subject, time, batch, sex, breed, age, cancer_type) |>
  kbl(col.names = c("Subject",
                    "Time",
                    "Batch",
                    "Sex",
                    "Breed",
                    "Age",
                    "Cancer Type"),
      row.names = FALSE,
      caption = "Sample Overview") |>
  collapse_rows(columns = 1, valign = "top") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), fixed_thead = T) |>
  scroll_box(width = "100%", height = "50%")
```

```{r}
mayo_abd <- read.table(mayo_gene_counts, header = TRUE, sep = "\t") |>
  dplyr::select("GeneID", "GeneName", "ORv.01__LN_FNA_Pre.TX_scaledTPM", "ORv.01__LN_FNA_Day_3_scaledTPM", "ORv.01__LN_FNA_Day_7_scaledTPM", "ORv.01__LN_FNA_Day_10_scaledTPM", "ORv.01__LN_FNA_Day_28_scaledTPM", "ORv.01__LN_FNA_Day_49_scaledTPM", "ORv.02__LN_FNA_Pre.TX_scaledTPM", "ORv.02__LN_FNA_Day_3_scaledTPM", "ORv.02__LN_FNA_Day_7_scaledTPM", "ORv.02__LN_FNA_Day_28_scaledTPM")|>
  filter(GeneName != "Unspecified") |>
  distinct(GeneName, .keep_all = TRUE) |>
  arrange(GeneName)
my_abd <- txi$counts |>
  as.data.frame() |>
  rownames_to_column("GeneID") |>
  left_join(tx2gene, by = c("GeneID" = "gene_id")) |>
  distinct(GeneID, .keep_all = TRUE) |>
  arrange(gene_name) |>
  filter(gene_name %in% mayo_abd$GeneName) |>
  dplyr::select(gene_name, starts_with("OR")) |>
  distinct(gene_name, .keep_all = TRUE) |>
  arrange(gene_name)

mayo_abd <- mayo_abd |>
  filter(GeneName %in% my_abd$gene_name) |>
  dplyr::select(-GeneID) |>
  arrange(GeneName)

i = 2
while (i < 12) {
  print(cor(mayo_abd[i], my_abd[i]))
  i = i+1
}
```

```{r echo = FALSE}
counts |>
  ggplot() +
  geom_histogram(aes(x = log2(count + 1)), position = "identity", bins = 200) +
    facet_grid(rows = vars(subject_lab),
             cols = vars(time_lab)) +
  theme_bw() +
  labs(x = "log2(count + 1)",
       y = "Number of Genes",
       title = "Count Distribution by Sample")
```

In order to fulfill the assumptions of the negative binomial model used by DESeq2, the distribution of counts should be approximately the same. Count data is expected to be right skewed.

# Differential Expression

## Model Fitting

This model uses the likelihood ratio test to model the change in gene expression over time, accounting for variation between subjects.

```{r}
dds <- run_model(txi,
                 sample_sheet,
                 design = ~ subject + time,
                 reduced = ~ subject,
                 test = "LRT",
                 min_count = min_count,
                 min_samples = min_samples)

df <- extract_coefs(dds, tx2gene, homologs, min_alpha = 1000000, min_logfold = 0) # unfiltered results

df_filter <- extract_coefs(dds, tx2gene, homologs, min_alpha = min_alpha, min_logfold = min_logfold) |>
  dplyr::filter(regulation != "nonDE" & coefficient != "subject_2_vs_1") # results filtered by alpha and log fold change

write.csv(df_filter, "diff_ex_genes.csv")
``` 

## Quality Control 

```{r}
counts_norm <- counts(dds, normalized = TRUE)  
counts_norm <- process_counts(counts_norm, sample_sheet, tx2gene)

counts_tbl <- counts |>
     dplyr::summarize(total_count = sum(count),
            mean_count = mean(count),
            sd_count = sd(count),
            max_count = max(count),
            num_genes = sum(count > 0),
            .by = c("subject", "time")) |>
    dplyr::mutate(normalization = "yes")
  
counts_norm_tbl <- counts_norm |>
     dplyr::summarize(total_count = sum(count),
            mean_count = mean(count),
            sd_count = sd(count),
            max_count = max(count),
            num_genes = sum(count > 0),
            .by = c("subject", "time")) |>
  dplyr::mutate(normalization = "no")
```

```{r echo = FALSE}
counts_tbl |>
  full_join(counts_norm_tbl) |>
  arrange(subject, time) |>
  kbl(col.names = c("Subject", "Time", "Total Count", "Mean Count", "Standard Deviation", "Max Count", "Number of Genes", "Normalized"),
      format.args = list(big.mark = ","),
      row.names = FALSE,
      caption = "Distribution of Gene Counts Across Samples") |>
  collapse_rows(columns = c(1,2), valign = "top") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), fixed_thead = T) |>
  scroll_box(width = "100%", height = "50%")
```

If there are large differences in the counts between samples, that could indicate problems with the sample or sequencing or uneven sequencing depth. Values are normalized using DESeq2. The normalized counts do NOT include genes that were removed during pre-filtering and DESeq2 internal filtering. 

```{r}
glimmaMDS(dds) # color by subject, shape by time.
# 46% of variance explained by subject
plotDispEsts(dds) # Black points should cluster around the fitted line.
```

```{r}
# visualizing shrinkage of log fold change
res <- results(dds)
plotMA(res, alpha = 0.1, ylim = c(-2, 2), cex = 1.5, main = "MA-plot before LFC shrinkage")

resLFC <- lfcShrink(dds, coef = "time_49_vs_0", type = "apeglm")
plotMA(resLFC, alpha = 0.1, ylim = c(-2, 2), cex = 1.5, main = "MA-plot after LFC shrinkage")
```


## Results
```{r}
glimmaMA(dds, groups = colData(dds)$time)
glimmaVolcano(dds, groups = colData(dds)$time)
```

```{r echo = FALSE}
df |>
  mutate(sig = padj < min_alpha & abs(log2FoldChange) > min_logfold) |>
  group_by(coefficient) |>
  summarise(DE_genes = sum(sig),
            Up = sum(sig & log2FoldChange > 0),
            Down = sum(sig & log2FoldChange < 0)) |>
  arrange(desc(DE_genes)) |>
   kbl(col.names = c("Coefficient", "Number of DE Genes", "Upregulated", "Downregulated"),
      format.args = list(big.mark = ","),
      row.names = FALSE,
      caption = "Number of Differentially Expressed Genes by Coefficient") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), fixed_thead = T) |>
  scroll_box(width = "100%", height = "50%")
```


```{r eval = FALSE}
up <- df_filter |>
  filter(regulation == "upregulated") |>
  pull(gene_id, coefficient) 
  
down <- df_filter |>
  filter(regulation == "downregulated") |>
  pull(gene_id) 

counts_norm |>
  filter(gene_id %in% up) |>
   left_join(homologs, by = c("gene_id" = "dog_gene_id")) |>
   dplyr::mutate(human_gene_name = coalesce(human_gene_name,
                                           gene_id)) |>
  ggplot(aes(x = time_num, y = count, shape = time, color = subject)) +
  geom_point() +
  #geom_smooth(method = "loess", se = FALSE) +
  theme_bw() +
  facet_wrap(~human_gene_name, scales = "free") +
  labs(
    x = "Time (days)",
    y = "Normalized count",
    title = "Upregulated Genes")

counts_norm |>
  filter(gene_id %in% down) |>
   left_join(homologs, by = c("gene_id" = "dog_gene_id")) |>
   dplyr::mutate(human_gene_name = coalesce(human_gene_name,
                                           gene_id)) |>
  ggplot(aes(x = time_num, y = count, shape = time, color = subject)) +
  geom_point() +
  #geom_smooth(method = "loess", se = FALSE) +
  theme_bw() +
  facet_wrap(~human_gene_name, scales = "free") +
  labs(
    x = "Time (days)",
    y = "Normalized count",
    title = "Downregulated Genes")
```

```{r eval = FALSE}
up3 <- df_filter |>
  filter(regulation == "upregulated" & coefficient == "time_3_vs_0") |>
  pull(gene_id) 

counts_norm |>
  filter(gene_id %in% up3) |>
   left_join(homologs, by = c("gene_id" = "dog_gene_id")) |>
   dplyr::mutate(human_gene_name = coalesce(human_gene_name,
                                           gene_id)) |>
  ggplot(aes(x = time_num, y = count, shape = time, color = subject)) +
  geom_point() +
  #geom_smooth(method = "loess", se = FALSE) +
  theme_bw() +
  facet_wrap(~human_gene_name, scales = "free") +
  labs(
    x = "Time (days)",
    y = "Normalized count",
    title = "Upregulated Genes")
```

```{r}
# create table here of gene and coefficients where it is differentially expressed
plot_list <- df_filter |>
  group_split(regulation, coefficient) |>
  map(function(df_sub) {
    gene_ids <- df_sub$gene_id
    counts_norm |>
      filter(gene_id %in% gene_ids) |>
      left_join(homologs, by = c("gene_id" = "dog_gene_id")) |>
      mutate(human_gene_name = coalesce(human_gene_name, gene_id)) |>
      ggplot(aes(x = time_num, y = count, shape = time, color = subject)) +
      geom_point() +
      theme_bw() +
      facet_wrap(~human_gene_name, scales = "free") +
      labs(title = paste(
          unique(df_sub$regulation),
          unique(df_sub$coefficient)))})
```

# Gene Set Enrichment
```{r}
msigdbr_collections()
# count how many human gene ids are NA

# gather pathway dataset
hallmark <- msigdbr(species = "Homo sapiens", collection = "H") |>
  dplyr::select(gs_name, ensembl_gene)
head(hallmark)

hallmark_terms <- msigdbr(species = "Homo sapiens", collection = "H") |>
  dplyr::select(gs_name, gs_description)

immune <- msigdbr(species = "Homo sapiens", collection = "C7", subcollection = "IMMUNESIGDB") |>
  dplyr::select(gs_name, ensembl_gene)

immune_terms <- msigdbr(species = "Homo sapiens", collection = "C7", subcollection = "IMMUNESIGDB") |>
  dplyr::select(gs_name, gs_description)

react <- msigdbr(species = "Homo sapiens", collection = "C2", subcollection = "CP:REACTOME") |>
  dplyr::select(gs_name, ensembl_gene)

react_terms <- msigdbr(species = "Homo sapiens", collection = "C2", subcollection = "CP:REACTOME") |>
  dplyr::select(gs_name, gs_description)

gobp <- msigdbr(species = "Homo sapiens", collection = "C5", subcollection = "GO:BP") |>
  dplyr::select(gs_name, ensembl_gene)

gobp_terms <- msigdbr(species = "Homo sapiens", collection = "C5", subcollection = "GO:BP") |>
  dplyr::select(gs_name, gs_description)

```

```{r}
df_hallmark <- run_gsea(dds, df, hallmark, hallmark_terms)
df_immune <- run_gsea(dds, df, immune, immune_terms)
df_react <- run_gsea(dds, df, react, react_terms)
df_gobp <- run_gsea(dds, df, gobp, gobp_terms)
```

```{r} 
df_hallmark |>
  mutate(pathway = str_sub(ID, 10),
         regulation = case_when(
    (NES > 0) ~ "upregulated",
    (NES < 0) ~ "downregulated",
    T ~ "nonDE")) |>
  ggplot(aes(x=reorder(pathway, -NES), y=NES, group = regulation)) +
  geom_bar(aes(fill = regulation), stat = "identity") +
  facet_wrap(~coefficient, scales = 'free_x') +
  theme_bw() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

```{r eval = FALSE}
plotCounts(dds, gene = 'ENSCAFG00805010402', intgroup = c('time'), main = 'CATSPERG')

counts_norm |>
  filter(gene_id == "ENSCAFG00805001673") |>
   left_join(homologs, by = c("gene_id" = "dog_gene_id")) |>
   dplyr::mutate(human_gene_name = coalesce(human_gene_name,
                                           gene_id)) |>
  ggplot(aes(x = time_num, y = count, shape = time, color = subject)) +
  geom_point() +
  theme_bw() +
  labs(
    x = "Time (days)",
    y = "Normalized count",
    title = "Expression")


# enrichplot::gseaplot2(ehall, geneSetID = c(1, 2, 3))
# GSEA_GO <- gseGO(geneList = all,
#                   ont = "BP",
#                   OrgDb = org.Hs.eg.db,
#                   keyType = "ENSEMBL",
#                   seed = TRUE)

comma_separated <- df_filter |>
  distinct(human_gene_name) |>
  pull(human_gene_name) |>
  paste(collapse = ", ")

print(comma_separated)
```

```{r}
sessionInfo()
```