---
title: "Longitudinal Bulk RNA-Seq Analysis of Canine Lymphoma Samples"
author: "Elise Hachfeld"
output: 
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
date: "2026-08-19"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE)
```

# Overview

add info here

# Setup

## Load Required Packages

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

#BiocManager::install(c("tidyverse",
#                       "stringr",
#                       "kableExtra",
#                       "rtracklayer",
#                       "tximport",
#                       "DESeq2",
#                       "Glimma",
#                       "apeglm",
#                       "msigdbr",
#                       "org.Hs.eg.db",
#                       "clusterProfiler"))

library(tidyverse)
library(stringr)
library(kableExtra)
library(rtracklayer)
library(tximport)
library(DESeq2)
library(Glimma)
library(apeglm)
library(msigdbr)
library(clusterProfiler)
library(org.Hs.eg.db)
```

## Helper Functions

### prep_tx2gene

Description:\
**Input:**\
**Output:**

```{r}
prep_tx2gene <- function(gtf_path) {
  print(paste0("Reading annotation file from ", gtf_path, "..."))
  tx2gene <- rtracklayer::readGFF(gtf_path) |>
    as.data.frame() |>
    dplyr::select(transcript_id,
                  gene_id,
                  gene_name,
                  gene_biotype) |>
    dplyr::filter(!is.na(transcript_id)) |>
    dplyr::distinct() |>
    dplyr::mutate(gene_name = coalesce(gene_name,
                                       gene_id)) # if no gene name present, use gene id
  print(paste0("Found ", length(unique(tx2gene$transcript_id)), " unique transcripts and ", length(unique(tx2gene$gene_id)), " unique genes."))
  return(tx2gene)
}
```

### prep_samples

Description:\
**Input:**\
**Output:**

```{r}
prep_samples <- function(sample_sheet_path) {
  print(paste0("Reading sample sheet from ", sample_sheet_path, "..."))
  sample_sheet <- read.csv(sample_sheet_path) |>
    dplyr::mutate(time_num = as.numeric(time),
                  time = factor(time, levels = sort(unique(time))),
                  time_lab = factor(paste0("Day ", time)),
                  subject = factor(subject, levels = sort(unique(subject))),
                  subject_lab = factor(paste0("Subject ", subject)),
                  sample_name = paste0("Subject ", subject, ", ", "Day ", time)) # readable sample name for visualizations
  rownames(sample_sheet) <- sample_sheet$sample_id
  print(paste0("Found ", nrow(sample_sheet), " samples in sample sheet."))
  return(sample_sheet)
}
```

### prep_kallisto

Description:\
**Input:**\
**Output:**

```{r}
prep_kallisto <- function(kallisto_path,
                          tx2gene,
                          sample_sheet,
                          transcripts = FALSE) {
  # NOTE: this will not include samples in the file path if they are not in the sample sheet
  print(paste0("Reading abundance.h5 files from ", kallisto_path, "..."))
  files <- file.path(kallisto_path, sample_sheet$sample_id, "abundance.h5") 
  print(paste0("Found ", length(files), " files in ", kallisto_path))
  names(files) <- sample_sheet$sample_id
  txi <- tximport(files,
                  type = "kallisto",
                  tx2gene = tx2gene,
                  txOut = transcripts, # change this to true if you want transcript level data
                  ignoreTxVersion = TRUE)
  return(txi)
}
```

### prep_homologs

Description:\
**Input:**\
**Output:**

```{r}
prep_homologs <- function(homolog_path) {
  print(paste0("Reading csv from ", homolog_path, "..."))
  homologs <- read.csv(homolog_path, sep = "\t") |>
    dplyr::filter(ref_species == "homo_sapiens") |>
    dplyr::select(ref_gene_stable_id,
                  ref_gene_name,
                  query_gene_stable_id,
                  query_gene_name) |>
    dplyr::rename(human_gene_id = ref_gene_stable_id,
                  human_gene_name = ref_gene_name,
                  dog_gene_id = query_gene_stable_id,
                  dog_gene_name = query_gene_name) |>
    group_by(dog_gene_id) |>
    summarise(across(everything(), ~paste0(unique(.), collapse = ", ")))
  print(paste0("Found ", length(unique(homologs$human_gene_id)), " homologous genes in humans."))
  return(homologs)
}
```

### run_model

Description:\
**Input:**\
**Output:**

```{r}
run_model <- function(txi = txi,
                      sample_sheet,
                      design,
                      reduced,
                      test = "Wald",
                      min_count,
                      min_samples) {
  dds <- DESeqDataSetFromTximport(txi,
                                  colData = sample_sheet,
                                  design = design)
  keep <- rowSums(counts(dds) >= min_count) >= min_samples # filtering for low counts
  dds <- dds[keep, ]

  if (test == "LRT") {
    dds <- DESeq(dds,
                 test = test,
                 reduced = reduced)
  } else {
    dds <- DESeq(dds,
                 test = test)
  }
  return(dds)
}
```

### process_counts

Description:\
**Input:**\
**Output:**

```{r}
process_counts <- function(counts, sample_sheet) {
  genes <- tx2gene |>
    dplyr::select(-transcript_id) |>
    dplyr::distinct()
  counts <- counts |>
    as.data.frame() |>
    rownames_to_column("gene_id") |>
    pivot_longer(cols = starts_with("OR"),
                 names_to = "sample_id",
                 values_to = "count") |>
    left_join(sample_sheet, by = "sample_id") |>
    left_join(genes, by = "gene_id") |>
    dplyr::distinct(gene_id, sample_id, .keep_all = TRUE)
  return(counts)
}
```

### extract_coefs

Description:\
**Input:**\
**Output:**

```{r}
extract_coefs <- function(dds,
                          tx2gene,
                          homologs,
                          min_alpha,
                          min_logfold) {
  # getting a data frame that has log fold change, p val and gene name for every coefficient
  coefficients <- resultsNames(dds)
  df_list <- list()
  for (coef in coefficients[-1]) {
    res <- lfcShrink(dds, coef = coef, type = "apeglm") |>
      as.data.frame() |>
      rownames_to_column("gene_id") |>
      left_join(tx2gene,
                by = "gene_id") |>
      dplyr::select(-transcript_id) |>
      dplyr::distinct(gene_id, .keep_all = TRUE) |>
      dplyr::filter(!is.na(padj) & padj < min_alpha & abs(log2FoldChange) > min_logfold) |> # genes with extreme outliers are set to NA in DESeq2
      dplyr::arrange(padj) |>
      dplyr::mutate(coefficient = coef)
      df_list <- append(df_list, list(res))
  }
  df <- do.call(rbind, df_list) |>
    left_join(homologs, by = c("gene_id" = "dog_gene_id")) |>
    mutate(regulation = case_when(
    (padj < min_alpha & abs(log2FoldChange) > min_logfold & log2FoldChange > 0) ~ "upregulated",
    (padj < min_alpha & abs(log2FoldChange) > min_logfold & log2FoldChange < 0) ~ "downregulated",
    T ~ "nonDE"))
  return(df)
}
```

### run_gsea

Description:\
**Input:**\
**Output:**

```{r}
run_gsea <- function(dds, main_df, path_data, path_terms){
  set.seed(42)
  df_list <- list()
  coefs <- resultsNames(dds)
  
  for (coef in coefs[-1]) {
    # create ranked list of genes
    ranked_genes <- main_df |>
      filter(coefficient == coef & !is.na(human_gene_name)) |>
      arrange(desc(log2FoldChange)) |>
      pull(log2FoldChange, name = human_gene_id)
  
    df_gsea <- GSEA(ranked_genes,
                  TERM2GENE = path_data,
                  TERM2NAME = path_terms,
                  seed = TRUE) |>
  as.data.frame() |>
  mutate(coefficient = coef)

  df_list <- append(df_list, list(df_gsea))
  }
  df_gsea <- do.call(rbind, df_list)
  return(df_gsea)
}
  
```


## Set Constants

### Parameters

```{r}
min_samples <- 2
min_count <- 10
min_alpha <- 0.05
min_logfold <- 0.5
```

### File Locations

```{r}
setwd("D:/canine-seq")
sample_sheet_path <- "D:/canine-seq/data/ORBIT/sample_sheet.csv"
gtf_path <- "D:/canine-seq/data/reference/Canis_lupus_familiarisgsd_UU_Cfam_GSD_1.0_116/Canis_lupus_familiarisgsd.UU_Cfam_GSD_1.0.116.gtf.gz"
kallisto_path <- "D:/canine-seq/runs/run_ORBIT_canfam4/results/kallisto"
homolog_path <- "D:/canine-seq/data/reference/Canis_lupus_familiarisgsd_UU_Cfam_GSD_1.0_116/Canis_lupus_familiaris-GCA_011100685.1-2021_03-homology.tsv"
```

## Data Import

```{r}
sample_sheet <- prep_samples(sample_sheet_path)
sample_sheet |>
  dplyr::select(subject, time, batch, sex, breed, age, cancer_type) |>
  kbl(col.names = c("Subject",
                    "Time",
                    "Batch",
                    "Sex",
                    "Breed",
                    "Age",
                    "Cancer Type"),
      row.names = FALSE,
      caption = "Sample Overview") |>
  collapse_rows(columns = 1, valign = "top") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), fixed_thead = T) |>
  scroll_box(width = "100%", height = "50%")

tx2gene <- prep_tx2gene(gtf_path)
homologs <- prep_homologs(homolog_path)
txi <- prep_kallisto(kallisto_path, tx2gene, sample_sheet)
counts <- process_counts(txi$counts, sample_sheet)
```

```{r}
counts |>
  ggplot() +
  geom_histogram(aes(x = log2(count + 1)), position = "identity", bins = 200) +
    facet_grid(rows = vars(subject_lab),
             cols = vars(time_lab)) +
  theme_bw() +
  labs(x = "log2(count + 1)",
       y = "Number of Genes",
       title = "Count Distribution by Sample")
```

In order to fulfill the assumptions of the negative binomial model used by DESeq2, the distribution of counts should be approximately the same. Count data is expected to be right skewed.

# Differential Expression

## Model Fitting

This model uses the likelihood ratio test to model the change in gene expression over time, accounting for variation between subjects.

```{r}
dds <- run_model(txi,
                 sample_sheet,
                 design = ~ subject + time,
                 reduced = ~ subject,
                 test = "LRT",
                 min_count = min_count,
                 min_samples = min_samples)

df <- extract_coefs(dds, tx2gene, homologs, min_alpha = 1000000, min_logfold = 0) # unfiltered results

df_filter_high_cutoff <- extract_coefs(dds, tx2gene, homologs, min_alpha = min_alpha, min_logfold = 1) |>
  dplyr::filter(regulation != "nonDE" & coefficient != "subject_2_vs_1") # results filtered by alpha and log fold change

df_filter <- extract_coefs(dds, tx2gene, homologs, min_alpha = min_alpha, min_logfold = min_logfold) |>
  dplyr::filter(regulation != "nonDE" & coefficient != "subject_2_vs_1") # results filtered by alpha and log fold change

write.csv(df_filter, "diff_ex_genes.csv")
``` 

## Quality Control 

```{r}
counts_norm <- counts(dds, normalized = TRUE)  
counts_norm <- process_counts(counts_norm, sample_sheet)

counts_tbl <- counts |>
     dplyr::summarize(total_count = sum(count),
            mean_count = mean(count),
            sd_count = sd(count),
            max_count = max(count),
            num_genes = sum(count > 0),
            .by = c("subject", "time")) |>
    dplyr::mutate(normalization = "yes")
  
counts_norm_tbl <- counts_norm |>
     dplyr::summarize(total_count = sum(count),
            mean_count = mean(count),
            sd_count = sd(count),
            max_count = max(count),
            num_genes = sum(count > 0),
            .by = c("subject", "time")) |>
  dplyr::mutate(normalization = "no")

counts_tbl |>
  full_join(counts_norm_tbl) |>
  arrange(subject, time) |>
  kbl(col.names = c("Subject", "Time", "Total Count", "Mean Count", "Standard Deviation", "Max Count", "Number of Genes", "Normalized"),
      format.args = list(big.mark = ","),
      row.names = FALSE,
      caption = "Distribution of Gene Counts Across Samples") |>
  collapse_rows(columns = c(1,2), valign = "top") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), fixed_thead = T) |>
  scroll_box(width = "100%", height = "50%")
```

If there are large differences in the counts between samples, that could indicate problems with the sample or sequencing or uneven sequencing depth. Values are normalized using DESeq2. The normalized counts do NOT include genes that were removed during pre-filtering and DESeq2 internal filtering. 

```{r}
glimmaMDS(dds) # color by subject, shape by time.
# 46% of variance explained by subject
plotDispEsts(dds) # Black points should cluster around the fitted line.
```

```{r}
# visualizing shrinkage of log fold change
res <- results(dds)
plotMA(res, alpha = 0.1, ylim = c(-2, 2), cex = 1.5, main = "MA-plot before LFC shrinkage")

resLFC <- lfcShrink(dds, coef = "time_49_vs_0", type = "apeglm")
plotMA(resLFC, alpha = 0.1, ylim = c(-2, 2), cex = 1.5, main = "MA-plot after LFC shrinkage")
```


## Results
```{r}
glimmaMA(dds, groups = colData(dds)$time)
glimmaVolcano(dds, groups = colData(dds)$time)
```

```{r}
df |>
  mutate(sig = padj < min_alpha & abs(log2FoldChange) > min_logfold) |>
  group_by(coefficient) |>
  summarise(DE_genes = sum(sig),
            Up = sum(sig & log2FoldChange > 0),
            Down = sum(sig & log2FoldChange < 0)) |>
  arrange(desc(DE_genes)) |>
   kbl(col.names = c("Coefficient", "Number of DE Genes", "Upregulated", "Downregulated"),
      format.args = list(big.mark = ","),
      row.names = FALSE,
      caption = "Number of Differentially Expressed Genes by Coefficient") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), fixed_thead = T) |>
  scroll_box(width = "100%", height = "50%")
```


```{r}
up <- df_filter |>
  filter(regulation == "upregulated") |>
  pull(gene_id, coefficient) 
  
down <- df_filter |>
  filter(regulation == "downregulated") |>
  pull(gene_id) 

counts_norm |>
  filter(gene_id %in% up) |>
   left_join(homologs, by = c("gene_id" = "dog_gene_id")) |>
   dplyr::mutate(human_gene_name = coalesce(human_gene_name,
                                           gene_id)) |>
  ggplot(aes(x = time_num, y = count, shape = time, color = subject)) +
  geom_point() +
  #geom_smooth(method = "loess", se = FALSE) +
  theme_bw() +
  facet_wrap(~human_gene_name, scales = "free") +
  labs(
    x = "Time (days)",
    y = "Normalized count",
    title = "Upregulated Genes")

counts_norm |>
  filter(gene_id %in% down) |>
   left_join(homologs, by = c("gene_id" = "dog_gene_id")) |>
   dplyr::mutate(human_gene_name = coalesce(human_gene_name,
                                           gene_id)) |>
  ggplot(aes(x = time_num, y = count, shape = time, color = subject)) +
  geom_point() +
  #geom_smooth(method = "loess", se = FALSE) +
  theme_bw() +
  facet_wrap(~human_gene_name, scales = "free") +
  labs(
    x = "Time (days)",
    y = "Normalized count",
    title = "Downregulated Genes")

```

```{r}
up3 <- df_filter |>
  filter(regulation == "upregulated" & coefficient == "time_3_vs_0") |>
  pull(gene_id) 

counts_norm |>
  filter(gene_id %in% up3) |>
   left_join(homologs, by = c("gene_id" = "dog_gene_id")) |>
   dplyr::mutate(human_gene_name = coalesce(human_gene_name,
                                           gene_id)) |>
  ggplot(aes(x = time_num, y = count, shape = time, color = subject)) +
  geom_point() +
  #geom_smooth(method = "loess", se = FALSE) +
  theme_bw() +
  facet_wrap(~human_gene_name, scales = "free") +
  labs(
    x = "Time (days)",
    y = "Normalized count",
    title = "Upregulated Genes")
```

```{r}
plot_list <- df_filter |>
  group_split(regulation, coefficient) |>
  map(function(df_sub) {
    gene_ids <- df_sub$gene_id
    counts_norm |>
      filter(gene_id %in% gene_ids) |>
      left_join(homologs, by = c("gene_id" = "dog_gene_id")) |>
      mutate(human_gene_name = coalesce(human_gene_name, gene_id)) |>
      ggplot(aes(x = time_num, y = count, shape = time, color = subject)) +
      geom_point() +
      theme_bw() +
      facet_wrap(~human_gene_name, scales = "free") +
      labs(title = paste(
          unique(df_sub$regulation),
          unique(df_sub$coefficient)))})
```

# Gene Set Enrichment
```{r}
msigdbr_collections()
# count how many human gene ids are NA

# gather pathway dataset
hallmark <- msigdbr(species = "Homo sapiens", collection = "H") |>
  dplyr::select(gs_name, ensembl_gene)
head(hallmark)

hallmark_terms <- msigdbr(species = "Homo sapiens", collection = "H") |>
  dplyr::select(gs_name, gs_description)

immune <- msigdbr(species = "Homo sapiens", collection = "C7", subcollection = "IMMUNESIGDB") |>
  dplyr::select(gs_name, ensembl_gene)

immune_terms <- msigdbr(species = "Homo sapiens", collection = "C7", subcollection = "IMMUNESIGDB") |>
  dplyr::select(gs_name, gs_description)

react <- msigdbr(species = "Homo sapiens", collection = "C2", subcollection = "CP:REACTOME") |>
  dplyr::select(gs_name, ensembl_gene)

react_terms <- msigdbr(species = "Homo sapiens", collection = "C2", subcollection = "CP:REACTOME") |>
  dplyr::select(gs_name, gs_description)

gobp <- msigdbr(species = "Homo sapiens", collection = "C5", subcollection = "GO:BP") |>
  dplyr::select(gs_name, ensembl_gene)

gobp_terms <- msigdbr(species = "Homo sapiens", collection = "C5", subcollection = "GO:BP") |>
  dplyr::select(gs_name, gs_description)

```

```{r}
df_hallmark <- run_gsea(dds, df, hallmark, hallmark_terms)
df_immune <- run_gsea(dds, df, immune, immune_terms)
df_react <- run_gsea(dds, df, react, react_terms)
df_gobp <- run_gsea(dds, df, gobp, gobp_terms)
```

```{r} 
df_hallmark |>
  mutate(pathway = str_sub(ID, 10),
         regulation = case_when(
    (NES > 0) ~ "upregulated",
    (NES < 0) ~ "downregulated",
    T ~ "nonDE")) |>
  ggplot(aes(x=reorder(pathway, -NES), y=NES, group = regulation)) +
  geom_bar(aes(fill = regulation), stat = "identity") +
  facet_wrap(~coefficient, scales = 'free_x') +
  theme_bw() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

```{r eval = FALSE}
plotCounts(dds, gene = 'ENSCAFG00805010402', intgroup = c('time'), main = 'CATSPERG')

counts_norm |>
  filter(gene_id == "ENSCAFG00805001673") |>
   left_join(homologs, by = c("gene_id" = "dog_gene_id")) |>
   dplyr::mutate(human_gene_name = coalesce(human_gene_name,
                                           gene_id)) |>
  ggplot(aes(x = time_num, y = count, shape = time, color = subject)) +
  geom_point() +
  theme_bw() +
  labs(
    x = "Time (days)",
    y = "Normalized count",
    title = "Expression")


# enrichplot::gseaplot2(ehall, geneSetID = c(1, 2, 3))
# GSEA_GO <- gseGO(geneList = all,
#                   ont = "BP",
#                   OrgDb = org.Hs.eg.db,
#                   keyType = "ENSEMBL",
#                   seed = TRUE)

comma_separated <- df_filter |>
  distinct(human_gene_name) |>
  pull(human_gene_name) |>
  paste(collapse = ", ")

print(comma_separated)
```

```{r}
sessionInfo()
```